---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>"
)
```

# StatistikkbankR

<!-- badges: start -->
<!-- badges: end -->

An R-package for accessing Statistikkbank data from Statistics Norway (SSB).

Statistics Norway (SSB) is Norway's central statistical authority.
It publishes hundreds of official statistics tables through its
Statistikkbank service, covering topics such as population,
economy, labour, housing, and health.

The `StatistikkbankR` package gives you a direct R interface to the
Statistikkbank API.  You can search the table catalogue, inspect
table structure, and download data as tidy `data.frame` or
tibble objects, directly in R.

## Installation

Install the development version from GitHub:

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("navikt/get_ssb_data")
```

## Finding a table

Use `ssb_search()` to find tables by keyword. The function returns
a data frame with one row per matching table; the `id` column is what
you will pass to other functions.

```{r search, eval = FALSE}
library(StatistikkbankR)

# Search for tables about population (in Norwegian)
results <- ssb_search("befolkning")
results[, c("id", "label", "updated")]

# Or search in English
ssb_search("population", language = "en")

# Retrieve all pages at once
all_results <- ssb_search("arbeid", fetch_all = TRUE)
nrow(all_results)
```

Alternatively, you can find tables of interest on SSB's [website](https://www.ssb.no/statbank).

## Inspecting a table before downloading

Once you have a table identifier, use `ssb_describe()` to see its
structure before downloading it.

```{r describe, eval = FALSE}
desc <- ssb_describe("07459")

# Table-level summary: title, last updated, estimated total rows
desc$table

# One row per dimension: id, label, role, number of codes
desc$dimensions
```

Use `ssb_codes()` to list the valid values you can use as filters
for a given dimension:

```{r codes, eval = FALSE}
# All valid codes for the Region dimension
ssb_codes("07459", "Region")

# All valid time periods
ssb_codes("07459", "Tid")
```

## Downloading data

### Without filters

Pass a table identifier to `get_ssb_data()` to download the full
table.  SSB enforces a cell limit on large tables; the function will
warn you and stop if the estimated row count is too high.

```{r download-all, eval = FALSE}
# Download all data (only works for small tables)
data <- get_ssb_data("05803")
head(data)
```

### Filtering by dimension

Pass dimension identifiers as named arguments to select only the
rows you need.  Each argument accepts a character vector of codes.

```{r download-filter, eval = FALSE}
# Download population for specific regions and the latest 5 years
pop <- get_ssb_data(
    "07459",
    Region = c("0301", "1103"),
    Tid    = "top(5)"
)
pop
```

### Server-side expressions

SSB supports a set of shorthand expressions that are evaluated on
the server before data is returned.  You can use these as filter
values:

| Expression | Meaning |
|---|---|
| `"top(n)"` | The n most recent time periods |
| `"bottom(n)"` | The n oldest time periods |
| `"all(*)"` | All codes (the default if you omit the argument) |

```{r expressions, eval = FALSE}
# The three most recent years for all regions
pop_recent <- get_ssb_data(
    "07459",
    Tid = "top(3)"
)
```

## Long format and singleton dimensions

By default `get_ssb_data()` returns data in wide format: one column
per measure code.  Use `table_format = "long"` to get a tidy long
table instead, with one row per observation.

```{r long, eval = FALSE}
pop_long <- get_ssb_data(
    "07459",
    Tid          = "top(3)",
    table_format = "long"
)
head(pop_long)
```

When filtering to a single code in one dimension, that dimension
collapses to a constant and is dropped from the wide output but
retained in long output by default.  Pass
`include_singleton_dims = TRUE` with `table_format = "long"` to
include those constant-value columns explicitly:

```{r singletons, eval = FALSE}
pop_one_region <- get_ssb_data(
    "07459",
    Region                 = "0301",
    Tid                    = "top(3)",
    table_format           = "long",
    include_singleton_dims = TRUE
)
```

## Codelists and geographic boundary changes

SSB provides aggregation codelists that let you request data at
a different grouping level — for example, historical municipality
boundaries — without manually remapping individual codes.

Use `ssb_codelists()` to discover which codelists are available:

```{r codelists, eval = FALSE}
ssb_codelists("07459")

# Filter to the geographic dimension
ssb_codelists("07459", dimension = "Region")
```

Use `ssb_codelist_details()` to inspect the group-to-member mapping
inside a specific codelist:

```{r codelist-details, eval = FALSE}
cl <- ssb_codelist_details("agg_KommSummer")
cl$codelist   # metadata
cl$values     # group codes and their members
```

Apply a codelist when downloading data by passing the `codelists`
and `output_values` arguments to `get_ssb_data()`:

```{r download-codelist, eval = FALSE}
pop_agg <- get_ssb_data(
    "07459",
    Tid           = "top(3)",
    codelists     = list(Region = "agg_KommSummer"),
    output_values = list(Region = "aggregated")
)
```

Or use the convenience wrapper `ssb_get_by_codelist()` which
identifies the correct dimension automatically:

```{r get-by-codelist, eval = FALSE}
pop_agg <- ssb_get_by_codelist(
    "07459",
    "agg_KommSummer",
    Tid          = "top(3)",
    output_value = "aggregated"
)
```

To join the aggregated result back to individual municipality codes,
use `ssb_expand_codelist_mapping()`:

```{r expand-codelist, eval = FALSE}
pop_expanded <- ssb_expand_codelist_mapping(
    data        = pop_agg,
    code_col    = "Region",
    codelist_id = "agg_KommSummer"
)
head(pop_expanded)
```

## Performance tips

### Caching

By default, `get_ssb_data()` and all helper functions cache table
metadata in memory for the duration of your R session.  Set
`cache = FALSE` to disable caching, or `refresh_metadata = TRUE`
to discard the cached copy and fetch fresh metadata:

```{r cache, eval = FALSE}
# Force a fresh metadata fetch (e.g. after SSB updates the table)
get_ssb_data("07459", Tid = "top(1)", refresh_metadata = TRUE)
```

### Large tables

SSB imposes a cell limit on queries.  `get_ssb_data()` estimates the
result size before sending the request and stops with a message if
the query is too large.  To proceed anyway, set
`override_large_query = TRUE`:

```{r large, eval = FALSE}
data <- get_ssb_data(
    "07459",
    override_large_query = TRUE
)
```
